\section*{\textsc{Curriculum Vitae}}

\section*{\textsc{Education}} \noindent\hrulefill \vspace{0.5em}
\begin{tabularx}{\textwidth}{@{}X r@{}}
\textbf{Doctor of Medicine (MD)}, \textsc{Stanford University} & \textit{Aug 2025 – Jun 2029} \\
\textbf{Doctor of Philosophy (PhD) in BME}, \textsc{Western University} & \textit{Sep 2021 – Aug 2025} \\
\textbf{Interdisciplinary Medical Sciences (BMSc)}, \textsc{Western University} & \textit{Sep 2017 – Jun 2021} \\
\end{tabularx}

\vspace{0.5em}
\section*{\textsc{Honors and Awards}} \noindent\hrulefill \vspace{0.5em}
% \begin{itemize}
  \item \textbf{Open Science Trailblazer Award}, Western University (Institutional, 2025) \hfill \textit{\$1,000}
  \item \textbf{Merit Abstract Award}, OHBM (International, 2024) \hfill \textit{\$1,000}
  \item \textbf{Parkinson Canada Graduate Student Award} (National, 2024) \hfill \textit{\$40,000}
  \item \textbf{NSERC CGS-D}, Canada Graduate Scholarship (National, 2023) \hfill \textit{\$120,000}
  \item \textbf{Mitacs Accelerate Fellowship} (National, 2022) \hfill \textit{\$25,000}
  \item \textbf{Ontario Graduate Scholarship} (Provincial, 2022 \& 2023) \hfill \textit{\$15,000 each}
  \item \textbf{Parkinson’s SWO Studentship} (Provincial, 2022) \hfill \textit{\$25,000}
  \item \textbf{Best Abstract}, Canadian Neuromodulation Conference (2022) \hfill \textit{\$1,500}
  \item \textbf{Best Platform Talk}, Clinical Neurological Research Day (2023) \hfill \textit{\$1,000}
  \item \textbf{Rising Youth Community Service Award} (National, 2020) \hfill \textit{\$5,000}

% \end{itemize}

\vspace{0.5em}
\section*{\textsc{Positions and Leadership}} \noindent\hrulefill \vspace{0.5em}
% \begin{itemize}
  \item \textbf{Co-Founder \& Board Member} — \href{https://webstraw.ca}{WebStraw}
  \item \textbf{Co-Founder \& Board Member} — \href{https://refuhope.com}{RefuHope}
  \item \textbf{Editor} — \href{https://stimulatingbrains.org}{Stimulating Brains Podcast}
  \item \textbf{Facilitator} — Deep Brain Stimulation Patient Support Group
  \item \textbf{Chair} — Biomedical Engineering Graduate Student Committee
  \item \textbf{Representative} — Schulich School of Medicine and Dentistry Committee
  \item \textbf{Representative} — Society of Graduate Students
  \item \textbf{Teaching Assistant} — Undergraduate Engineering (Years 1–2)
% \end{itemize}

\vspace{0.5em}
\section*{\textsc{Peer-Reviewed Publications}} \noindent\hrulefill \vspace{0.5em}
% \begin{enumerate}
  \item Abbass M., Taha A., Gilmore G., et al. (2024). \textit{Quantifying localization and registration accuracy using anatomical fiducials: application to deep brain stimulation}. \textit{Imaging Neuroscience}, In press.
  \item Haast R.M., Kai J., Taha A., et al. (2024). \textit{Mapping the topographic organization of the human zona incerta using diffusion MRI}. \textit{eLife}, 14:RP103530.
  \item Ogunsanya F., Taha A., et al. (2023). \textit{MRI-degad: Towards accurate conversion of gadolinium-enhanced MRI to non-contrast-enhanced scans via CNNs}. \textit{Int J CARS}, 19, 1469.
  \item Taha A., et al. (2023). \textit{Magnetic resonance imaging datasets with anatomical fiducials for quality control and registration}. \textit{Sci Data}, 10, 449.
  \item Taha A., Gilmore G., Khan A., Lau J.C. (2022). \textit{An indirect deep brain stimulation targeting tool using salient anatomical fiducials}. \textit{Neuromodulation}, 25(6).
  \item Abbass M., Gilmore G., Taha A., et al. (2022). \textit{Application of anatomical fiducials to a clinical dataset of Parkinson’s disease patients}. \textit{Brain Struct Funct}, 227, 393–405.
  \item Elias G.J.B., Germann J., Loh A., Boutet A., Taha A., et al. (2022). \textit{Normative connectomes and their use in DBS}. In A. Horn (Ed.), \textit{Connectomic Deep Brain Stimulation} (pp. 245–274). Academic Press.
  \item Germann J., Mameli M., Elias G.J.B., Loh A., Taha A., et al. (2021). \textit{Deep brain stimulation of the habenula: Systematic review and clinical trial registries}. \textit{Front Psychiatry}, 12, 730931.
  \item Boutet A., Loh A., Chow C., Taha A., et al. (2021). \textit{Advancements in MRI sequences for visualizing functional neurosurgical targets}. \textit{J Neurosurg}, 1–14.
% \end{enumerate}
\vspace{0.5em}
\section*{\textsc{Select Conference Presentations}} \noindent\hrulefill \vspace{0.5em}
% \begin{enumerate}
    \item Taha A. et al. (2025). AutoAFIDs: BIDS App for landmark detection, QC, stereotactic targeting and brain charting. \textit{OHBM}, Oral (International).
    \item Taha A. et al. (2024). Bridging brain coordinates and machine learning for surgical targeting and morphometric mapping. \textit{OHBM}, Oral (International).
    \item Taha A. et al. (2023). Mapping the subcortical connectome in Parkinson’s disease using 7-Tesla diffusion MRI. \textit{OHBM}, Poster (International).
    \item Taha A. et al. (2023). Open software infrastructure for automatic brain landmark localization for neuroimaging applications. \textit{OHBM}, Poster (International).
    \item Taha A. et al. (2023). Localizing the subthalamic nucleus using validated anatomical landmarks and machine learning. \textit{OHBM}, Poster (International).
    \item Kai J., Haast R.A.M., Taha A. et al. (2023). Cortical connections of the zona incerta revealed by in vivo diffusion. \textit{OHBM}, Poster (International).
    \item Gilmore G., Taha A. et al. (2022). Retrospective analysis of factors impacting SEEG accuracy at a Canadian center. \textit{American Epilepsy Society}, Poster (International).
    \item Taha A. et al. (2022). Deep brain stimulation targeting using an open-access anatomical fiducial framework. \textit{Imaging Network Ontario}, Oral (Provincial).
    \item Taha A. et al. (2022). Localizing the subthalamic nucleus on T1w gadolinium-enhanced imaging. \textit{Neuroscience Research Day}, Oral (Provincial).
% \end{enumerate}
